The ChEMBL database as Linked Open Data

### 2012-07-09 article.tex
More tuning

In this study we extended Bioclipse DS with remote access to ChEMBL-RDF and ChemSpider. 
This enables users browsing chemical structures in Bioclipse to look up near neighbors in ChemSpider 
via the ChemSpider Web API (SOAP), and for the found compounds to query ChEMBL-RDF for 
available interaction data. The results are presented alongside predictive models in Bioclipse (see Fig \ref{fig:bioclipse-ds}, and can be used for decision support when evaluating chemical structures and consider 
strategies for optimization.

@@

Designing a molecule, which is selective to one target over another will often be considered a 
successful outcome in a drug design process. If the target in question is a protein, it is easy
to understand that as sequence identity amongst residues, which contribute to potential small 
molecule binding sites remains high so does the issue of selectivity. That is, the molecule being 
designed may also bind to the equivalent binding site in the  closely related protein 
target and may lead to undesirable consequences. The ChEMBL data model links molecules to 
targets using different activity types recorded in the literature. Using activity types, 
which act as a measure of binding affinity, such as IC50 or Ki and applying activity value cuts
offs it is possible to identify molecules which have higher binding affinity to certain targets 
compared to others. Taking this one step further it is possible to identify a set molecules, 
which have a high affinity to protein A (e.g. IC50 value < 50nM) and low affinity to protein 
B (e.g. IC50 value > 200nM).

@@

\begin{tiny}
\begin{verbatim}
select distinct ?molecule WHERE 
{ 
  ?assay ?p chembl:Assay ;
  chembl:hasTarget ?target .
  ?activity ?p chembl:Activity ;
  chembl:forMolecule ?molecule ;
  chembl:onAssay ?assay ;
  chembl:type ?type ;
  chembl:standardUnits ?unit ;
  chembl:standardValue ?value;
  chembl:relation ?relation
  FILTER (
    ?type = "IC50"  &&
    ?unit = "nM"    &&
    ?value  < 50    &&
    ?relation = "="
  ) .
  ?target ?p chembl:Target;
  dc:identifier ?id
  FILTER (
    ?id = "uniprot:P24941"
  ) .
  { 
    select ?molecule WHERE 
    {
      ?assay ?p chembl:Assay ;
      chembl:hasTarget ?target .
      ?activity ?p chembl:Activity ;
      chembl:forMolecule ?molecule ;
      chembl:onAssay ?assay;
      chembl:type ?type ;
      chembl:standardUnits ?unit ;
      chembl:standardValue ?value ;
      chembl:relation ?relation
      FILTER (
        ?type = "IC50"  &&
        ?unit = "nM"    &&
        ?value  > 200   &&
        ?relation = "="
      ) .
      ?target ?p chembl:Target;
      dc:identifier ?id
      FILTER (
        ?id = "uniprot:P11802"
      ) .
    }
  }
}
\end{verbatim}
\end{tiny}

>>>>

In this study we extended Bioclipse DS with remote access to ChEMBL-RDF and ChemSpider. 
This enables users browsing chemical structures in Bioclipse to look up near neighbors in ChemSpider 
via the ChemSpider Web API (SOAP), and for the found compounds to query ChEMBL-RDF for 
available interaction data. The results are presented alongside predictive models in Bioclipse
(see Fig~\ref{fig:bioclipse-ds}), and can be used for decision support when evaluating chemical
structures and consider strategies for optimization.

@@

Designing a molecule, which is selective to one target over another will often be considered a 
successful outcome in a drug design process. If the target in question is a protein, it is easy
to understand that as sequence identity amongst residues, which contribute to potential small 
molecule binding sites remains high so does the issue of selectivity. That is, the molecule being 
designed may also bind to the equivalent binding site in the  closely related protein 
target and may lead to undesirable consequences. The ChEMBL data model links molecules to 
targets using different activity types recorded in the literature. Using activity types, 
which act as a measure of binding affinity, such as IC$_{50}$ or K$_{i}$ and applying activity value cuts
offs it is possible to identify molecules which have higher binding affinity to certain targets 
compared to others. Taking this one step further it is possible to identify a set molecules, 
which have a high affinity to protein A (e.g. IC$_{50}$ value $<$ 50~nM) and low affinity to protein 
B (e.g. IC$_{50}$ value $>$ 200~nM).

@@

\begin{small}
\begin{verbatim}
select distinct ?molecule WHERE 
{ 
  ?assay ?p chembl:Assay ;
  chembl:hasTarget ?target .
  ?activity ?p chembl:Activity ;
  chembl:forMolecule ?molecule ;
  chembl:onAssay ?assay ;
  chembl:type ?type ;
  chembl:standardUnits ?unit ;
  chembl:standardValue ?value;
  chembl:relation ?relation
  FILTER (
    ?type = "IC50"  &&
    ?unit = "nM"    &&
    ?value  < 50    &&
    ?relation = "="
  ) .
  ?target ?p chembl:Target;
  dc:identifier ?id
  FILTER (
    ?id = "uniprot:P24941"
  ) .
  { 
    select ?molecule WHERE 
    {
      ?assay ?p chembl:Assay ;
      chembl:hasTarget ?target .
      ?activity ?p chembl:Activity ;
      chembl:forMolecule ?molecule ;
      chembl:onAssay ?assay;
      chembl:type ?type ;
      chembl:standardUnits ?unit ;
      chembl:standardValue ?value ;
      chembl:relation ?relation
      FILTER (
        ?type = "IC50"  &&
        ?unit = "nM"    &&
        ?value  > 200   &&
        ?relation = "="
      ) .
      ?target ?p chembl:Target;
      dc:identifier ?id
      FILTER (
        ?id = "uniprot:P11802"
      ) .
    }
  }
}
\end{verbatim}
\end{small}
